---
title: "RUV-seq on polar infection"
output: 
  html_notebook: 
editor_options:
  chunk_output_type: inline
---

```{r include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(include = TRUE)
library(tidyverse)
library(greekLetters)
library(RUVSeq)
library(EDASeq)
library(DESeq2)
library("ggvenn")
library(ggprism)
library(eulerr)
library(rcartocolor)
theme_set(theme_prism(base_size = 14))
```




```{r}
counts <- read.table(here::here("long_infection_human_reads/long_RNA_library/data/all_reads_STAR"), header = F, sep = "\t", stringsAsFactors = F)
#Next manipulate the table to remove redundant columns and only keep relevant columns (1 columns with gene names, and the rest with counts for reverse stranded library)
counts_simple <- counts %>% 
  dplyr::select(Gene_ID = V1, UI92 = V4, UI93 = V8, UI96 = V12, N92 = V16, N93 = V20, WT96 = V24, WT91 = V28, WT94 = V32, N95 = V36)
#Filter really low expressed counts out
filter <- apply(counts_simple [,2:10],1,function(x) mean(x)>5)
counts_simple <- counts_simple [filter,]
```

```{r edaseq, fig.width=8, fig.height=6}
#Use EDAseq to normalize the libraries
c_mat <- counts_simple %>% 
                     dplyr::select(-where(is.character)) %>% 
                     dplyr::slice(-1:-4) %>% 
                     as.matrix()
rownames(c_mat) <- counts_simple$Gene_ID [-1:-4]
pData <- data.frame(genotype=vector(mode = "logical", length = ncol(c_mat)),
                    timepoint = vector(mode = "logical", length = ncol(c_mat)),
                    samples = colnames(c_mat),
                    row.names = colnames(c_mat))
pData <- pData %>% 
  mutate(genotype = case_when(
    str_detect(samples, "WT") ~ "WT",
    str_detect(samples, "UI") ~ "UI",
    str_detect(samples, "N") ~ "N"
  ), timepoint = 9) %>%
  dplyr::select(-samples)
pData$group <- factor(paste0(pData$genotype,pData$timepoint))

data <- newSeqExpressionSet(counts=c_mat,
                            phenoData=pData)
#Normalize the data next
dataNorm <- betweenLaneNormalization(data, which="upper") 
#For DE we will create this analysis with offset values
dataOffset <- betweenLaneNormalization(data,
                                       which="upper",offset=TRUE)
#Plot the counts 
#pdf(here::here("long_infection_human_reads/long_RNA_library/results/final RUV figures/EDAseq_normalization.pdf"), width = 8, height = 6)
pal_choice <- "Bold"
pData <- pData %>% 
  mutate(graph_cols  = case_when(
    genotype == "UI" ~ carto_pal(12,pal_choice) [10],
    genotype == "WT" ~ carto_pal(12,pal_choice) [2],
    genotype == "N" ~ carto_pal(12,pal_choice) [1]
  ))
plotPCA(data, col = pData$graph_cols, main = "Unnormalized counts")
plotPCA(dataOffset, col = pData$graph_cols, main = "Normalized counts")
plotRLE(data, col = pData$graph_cols, main = "RLE Unnormalized counts")
plotRLE(dataOffset, col = pData$graph_cols, main = "RLE Normalized counts")
boxplot(data, main = "Unnormalized counts", col = pData$graph_cols)
boxplot(dataNorm, main = "Normalized counts", col = pData$graph_cols)
#dev.off()
```

Try RUV-Seq. First confirm that library size normalization helps somewhat with data, but not enough for clear separation of genotypes

Prepare the RUV-seq objects and perform RUVs normalization
``` {r echo=TRUE, results = 'asis', include=TRUE, fig.width=8, fig.height=6}
scidx <- matrix(data = c(str_subset(rownames(pData), "UI"),
                         str_subset(rownames(pData), "WT"),
                         str_subset(rownames(pData), "N")),  nrow = 3, byrow = T)
cidx <- rownames(counts(data))
rv1 <- RUVs(dataOffset, cidx, k=1, scidx, epsilon = 1, isLog = F)
rv2 <- RUVs(dataOffset, cidx, k=2, scidx, epsilon = 1, isLog = F)
#pdf(here::here("long_infection_human_reads/long_RNA_library/results/final RUV figures/Normalization effect on the library and RUV normalization.pdf"), width = 8, height = 6)
plotPCA(dataOffset, col = pData$graph_cols, main = "UQ normalized counts")
plotPCA(rv1, col = pData$graph_cols, main = "UQ normalized counts + RUVs with 1 k")
plotPCA(rv2, col = pData$graph_cols, main = "UQ normalized counts + RUVs with 2 k")
plotRLE(dataOffset, col = pData$graph_cols, main = "UQ normalized counts")
plotRLE(rv1, col = pData$graph_cols, main = "UQ normalized counts + RUVs with 1 k")
plotRLE(rv2, col = pData$graph_cols, main = "UQ normalized counts + RUVs with 2 k")
#dev.off()
```
#The RUV normalization is done, now we have normalization factors or weights, that can be fed into DESeq2
Keep an eye on this 
https://bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html#using-ruv-with-deseq2


``` {r echo=TRUE, results = 'asis', include=TRUE}
dds <- DESeqDataSetFromMatrix(countData = counts(dataOffset),
                              colData = pData,
                              design = ~  group)
normFactors <- exp(-1 * offst(dataOffset))
normFactors <- normFactors / exp(rowMeans(log(normFactors)))
normalizationFactors(dds) <- normFactors
dds <- DESeq(dds)
rld <- vst(dds)
DESeq2::plotPCA(rld, intgroup = "group")
resWTvsUI <- results(dds, contrast = c("group","WT9", "UI9"))
resWTvsN <- results(dds, contrast = c("group","WT9", "N9"))
resNvsUI <- results(dds, contrast = c("group","N9", "UI9"))
map(list(resWTvsUI,resWTvsN,resNvsUI),summary)


dds1 <- DESeqDataSetFromMatrix(countData = counts(rv1),
                              colData = pData(rv1),
                              design = ~ W_1 + group)
normFactors <- exp(-1 * offst(dataOffset))
normFactors <- normFactors / exp(rowMeans(log(normFactors)))
normalizationFactors(dds1) <- normFactors
dds1 <- DESeq(dds1)
rld1 <- vst(dds1)
DESeq2::plotPCA(rld1, returnData = F, intgroup = "group")

mat <- limma::removeBatchEffect(assay(rld1),
                                covariates = rld1$W_1, design = model.matrix(~ genotype, colData(rld1)))
rld1.nobatch <- rld1
assay(rld1.nobatch) <- mat
DESeq2::plotPCA(rld1.nobatch, returnData = F, intgroup = "group")

resWTvsUI.1 <- results(dds1, contrast = c("group","WT9", "UI9"))
resWTvsN.1 <- results(dds1, contrast = c("group","WT9", "N9"))
resNvsUI.1 <- results(dds1, contrast = c("group","N9", "UI9"))
map(list(resWTvsUI.1,resWTvsN.1,resNvsUI.1),summary)

#Now repeat with 2 factors accounted for:
dds2 <- DESeqDataSetFromMatrix(countData = counts(rv2),
                              colData = pData(rv2),
                              design = ~ W_1 + W_2 + group)
normalizationFactors(dds2) <- normFactors
dds2 <- DESeq(dds2)
rld2 <- vst(dds2)
DESeq2::plotPCA(rld2, returnData = F, intgroup = "group")

mat <- limma::removeBatchEffect(assay(rld2),
                                covariates = matrix(data = c(rld2$W_1,rld2$W_2), ncol = 2),
                                design = model.matrix(~ genotype, colData(rld2)))
rld2.nobatch <- rld2
assay(rld2.nobatch) <- mat
DESeq2::plotPCA(rld2.nobatch, returnData = F, intgroup = "group")

resWTvsUI.2 <- results(dds2, contrast = c("group","WT9", "UI9"))
resWTvsN.2 <- results(dds2, contrast = c("group","WT9", "N9"))
resNvsUI.2 <- results(dds2, contrast = c("group","N9", "UI9"))
walk(list(resWTvsUI.2,resWTvsN.2,resNvsUI.2),summary)
```

Generally, the differences between 1 and 2 factors are not that large. Check how differentially expressed genes behave
```{r eulerplot, fig.width=6, fig.height=4}
wtvsn.res <- list(resWTvsN, resWTvsN.1,resWTvsN.2) %>% 
  map(as.data.frame) %>%  
  map(~(filter(.,padj < 0.05))) %>% 
  map(function(x) rownames(x))
names(wtvsn.res) <- c("No RUV","RUVs k=1","RUVs k=2")
#Visualize these results on a Venn diagram
#pdf(here::here("long_infection_human_reads/long_RNA_library/results/final RUV figures/RUV effect on DE genes between WT and EscN infection.pdf"), width = 6, height = 4)
eulerplot <- euler(wtvsn.res, shape = "circle")
plot(eulerplot, quantities = list(T, type = c("counts"), fontsize = 12), 
     fills = c(carto_pal(7,"DarkMint") [1],carto_pal(7,"DarkMint") [3], carto_pal(7,"DarkMint") [7]),
     adjust_labels = T, alpha = 0.7, legend = T, labels = F)
plot(eulerplot, quantities = list(T, type = c("counts", "percent"), fontsize = 12), 
     fills = c(carto_pal(7,"DarkMint") [1],carto_pal(7,"DarkMint") [3], carto_pal(7,"DarkMint") [7]),
     adjust_labels = T, alpha = 0.7, legend = T, labels = F)
#dev.off()
```

Check whether 1 factor is enough to get noticeable correlation between genotype and PC. 
```{r, fig.width=8, fig.height=6}
library(PCAtools)
p1 <- pca(assay(rld1.nobatch),colData(dds1), removeVar = 0.1)
p2 <- pca(assay(rld2.nobatch),colData(dds2), removeVar = 0.1)
#pdf(here::here("final RUV figures/PCAtools outputs with RUV 1 or 2.pdf"), width = 12, height = 10)
biplot(p1, colby = "genotype", shape = "genotype",gridlines.major = FALSE, title = "RUVs k=1")
biplot(p2, colby = "genotype", shape = "genotype",gridlines.major = FALSE, title = "RUVs k=2")
eigencorplot(p1,
             components = getComponents(p1,1:9),
    metavars = c('group','genotype', 'W_1'),
    col = c('white', 'cornsilk1', 'gold', 'forestgreen', 'darkgreen'),
    cexCorval = 1.2,
    fontCorval = 2,
    posLab = 'all',
    rotLabX = 45,
    scale = TRUE,
    main = bquote(PC ~ "RUVs k=1" ~ Pearson ~ r^2 ),
    plotRsquared = TRUE,
    corFUN = 'pearson',
    corUSE = 'pairwise.complete.obs',
    corMultipleTestCorrection = 'BH',
    signifSymbols = c('****', '***', '**', '*', ''),
    signifCutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1))
eigencorplot(p2,
             components = getComponents(p2,1:9),
    metavars = c('group','genotype', 'W_1','W_2'),
    col = c('white', 'cornsilk1', 'gold', 'forestgreen', 'darkgreen'),
    cexCorval = 1.2,
    fontCorval = 2,
    posLab = 'all',
    rotLabX = 45,
    scale = TRUE,
    main = bquote(PC ~ "RUVs k=2" ~ Pearson ~ r^2 ),
    plotRsquared = TRUE,
    corFUN = 'pearson',
    corUSE = 'pairwise.complete.obs',
    corMultipleTestCorrection = 'BH',
    signifSymbols = c('****', '***', '**', '*', ''),
    signifCutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1))
#dev.off()
```
From this it is clear that 2 unwanted factors is better than 1. We get majority of the same DE genes as non-adjusted or 1 factor adjusted objects.
